---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# grass

<!-- badges: start -->
<!-- badges: end -->

grass (Genetic Regression Approximation through Summary Statistics) is an in-development R package to describe various regression models using only genome-wide association study (GWAS) summary statistics.

Currently, grass supports the linear modeling of complex phenotypes defined via functions of other phenotypes.
Supported functions include:

* linear combinations (E.g. $\phi_1y_1 + \phi_2y_2$)
* products (E.g. $y_1\circ y_2$)
* logical combinations (E.g. $y_1\wedge y_2$ or $y_1\vee y_2$)

## Installation

grass is not currently available on CRAN.

You can install the in-development version of grass from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("jackmwolf/grass")
```
## Example

Let's model the first principal component of $y_1, y_2$ and $y_3$ using summary statistics.

```{r}
library(grass)
dat <- grass::cont_data[c("g", "x", "y1", "y2", "y3")]
head(dat)
```

First, we need our assumed summary statistics. 

```{r}
means <- colMeans(dat)
covs  <- cov(dat)
n     <- nrow(dat)
```

In addition, we need our weights, the first principal component vector of the responses' covariance matrix.

```{r}
SigmaY <- covs[c("y1", "y2", "y3"), c("y1", "y2", "y3")]
phi <- eigen(SigmaY)$vectors[, 1]
```

Then, we can calculate the linear model by using `calculate_lm_combo`

```{r}
model_pcss <- calculate_lm_combo(
  means = means, covs = covs, n = n, phi = phi, add_intercept = TRUE
  )
model_pcss
```

Here's the same model using individual patient data. 

```{r}
pc_1 <- prcomp(x = dat[c("y1", "y2", "y3")])$x[, "PC1"]
mod_ipd <- lm(pc_1 ~ 1 + g + x, data = dat)
summary(mod_ipd)$coef
```

The difference in test statistic estimates can be attributed to the non-uniqueness of optimal principal component vectors by multiplication by $-1$.


## References
Following are the key references for the functions in this package

* Wolf, J.M., Barnard, M., Xueting, X., Ryder, N., Westra, J., and Tintle, N. 
  (2020). Computationally efficient, exact, covariate-adjusted genetic principal
  component analysis by leveraging individual marker summary statistics from 
  large biobanks. *Pacific Symposium on Biocomputing*, 25, 719-730. 
  [https://doi.org/10.1142/9789811215636_0063](https://doi.org/10.1142/9789811215636_0063).
  
* Gasdaska A., Friend D., Chen R., Westra J., Zawistowski M., Lindsey W. and 
  Tintle N. (2019) Leveraging summary statistics to make inferences about 
  complex phenotypes in large biobanks. *Pacific Symposium on Biocomputing*, 24, 
  391-402.
  [https://doi.org/10.1142/9789813279827_0036](https://doi.org/10.1142/9789813279827_0036).
